Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-02-24 manuscript/cpcrw_manuscript.Rmd
Updated with comments from @apeyton

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Such laboratory studies have weaknesses, but their controllled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that soils maintained at field moisture and high temperature would lose more C, both as CO~2~ and CH~4~, than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity. In addition, cores subject to drought (reduced soil moisture) would have a higher potential to lose more ‘complex’ carbon compounds (aromatic C) as these compounds are desorbed due to increased surface tension. TODO: maybe change/reword this last sentence.

@@

@@

@@

@@

This research was supported by Office of Science of the U.S. Department of Energy as part of the Terrestrial Ecosystem Sciences Program. The Pacific Northwest National Laboratory is operated for DOE by Battelle Memorial Institute under contract DE-AC05-76RL01830.

@@

@@

@@

>>>>

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Such laboratory studies have weaknesses, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}. TODO: add Kirschbaum 2006. Most studies have focused on surface soils or permafrost soils, however, neglecting deep soils that experience fluctuating temperature with season and moisture with potential permafrost thaw ..or something less awkward (TODO; CITE). Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that soils maintained at field moisture and high temperature would lose more C, both as CO~2~ and CH~4~, than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity.

@@

APS comments to address re above - TODO:

* It doesn’t look like there was an interactive effect of temp*moisture, but just checking if you included that in the stat model?
* This is good IF the controlled drought was really following the same dry-down rate of the 4C drought treatment (see comment on Figure 2 below). This way a comparison between 20C drought and 4C drought is more appropriate and meaningful.
* Totally surprised by this [last sentence above]. Assumed CH4 would be higher in cores sitting in water (i.e. field moist) – perhaps this CH4→ CO2 ‘effect’ is occurring in these cores as well!?!! 
* Cumulative CO2 between cores held at 4˚ C and cores held at 20˚ C in the“field moisture” treatment did not (???!!!) differ (TRUE?!! Kinda looks that way and that is interesting and useful

@@

Treat et al. 2014 also showed that CH4 production from organic permafrost soils has a positive relationship with temperature (cites several studies: Lupasco et al. 2012, etc.).  perhaps mention (briefly) here - that these are not surface layer soils (where majority of microbial activity and C mineralization of labile C takes place) and that taking them out of depth (where they are not exposed to O2 as readily, etc.) for our laboratory incubation may differ from in situ CO2/CH4 production.

Increased temperature alone may not drive C mineralization rates in forest-mineral soils (Giardina and Ryan, 2000), but its corresponding increase in soil moisture content can lead to changes in microbial community structure and enhance anaerobic activity that may amplify CO2, CH4 and N2O fluxes (Xue et al, 2016).

@@

"Barfed bullets" from APS:

*	Seeta’s paper: Mineral soil: greenhouse treatment (warming treatment) = 0.646 ± 0.04 GREATER THAN control = 0.43 ± 0.05 C-CO2 per g soil per day in 14 day incubation ( α ≤ 0.1). Mineral soil also had greater active biomass and extractable-OM). (Sistla et al. 2013 Nature).  
*	Xue et al. 2016: active layer, contained organic horizon: soil moisture increased in warmed plots: ~50% moisture compared to ~45% (p < 0.05. C Flux (g/m2) from ecosystem respiration was greater in warmed plots (~500 g/m2) compared to control plots (~400) (p < 0.05). Authors remarked that soil moisture/climate was important factor in shifting functional composition of microbial community (GeoChip), esp. with anaerobic activity. We did not see differences in our cum C-CH4 loss with temperature or moisture, but we did see relationship between moisture and CH4 in both field moist treatments (4˚ and 20˚) and 20˚ drought. 
*	Treat et al. 2015. Meta-analysis of anaerobic incubations.  Mineral soils 20-100 cm depth had a mean CO2 production (ug CO2-C /g C/day) of 100.6 +/- 10.3 (33.2 +/- 3.7 per g DW)  (7840 per m2/d). All soils < 10C were 67 +/- 16 and > 20C was 172 +/- 17. Dry soils 68 +/- 8 and Fluctuating = 364 +/- 60 and inundated = 105 +/- 11. 
*	Schaufler et al. 2010: 33.7 +/- 2.7C02-C (mg C/ m2/ hr) averaged across all temperature (5-20C) and moisture (20-80% WFPS) in forested, boreal site, with 33.7 +/- 2.7.  CH4-C = -30.2 +/- 4.5. Optimum moisture content was 40% WFPS for this site – or highest CO2 evolved at this moisture content. Carbon dioxide emissions were linearly correlated with temperature (p = 0.000) and exponentially correlated with WFPS (p = 0.000). CH4 uptake (oxidation) was highest in the northern sites (included boreal site), and was also the driest soil moisture conditions. -44.8 +/- 5.5. ug CH4-C/m2/hr and decreased with increasing soil moisture.  Effect of WFPS on CH4 oxidation, exponential, p  = 0.000.  
*	Wu et al. 2010. Mongolian steppe 900-1400msl. Soils = chernozems, but there was a gleyosol. Soil moisture and temperature influenced CH4. Soil drought limited CH4 uptake at all 15C. CO2 in meadow sites – increased with temperature (< 30 mg C/m2/hr @ 5C to 90 at 15C) and with moisture (<30-50 in dry treatment from 5-15C, and 50-90 in high rain treatment at 5-15C).

Why would we expect to see greater CO2 evolved in higher temps?

*	Increased temp (to a certain degree) increases enzyme kinetics (i.e. Arrhenius equation) – rate of enzyme reactions increase, thus decomposition increases (and CO2 production increases) (Kirschbaun  1995, 2006 for SOM-temp, Wallenstein/Allison/Schimel for enzyme-temp, etc. so many others)
*	Increased temperature enhances the metabolic rate of microorganisms – thus increasing the production of extracellular enzymes (state somewhere prior that extracellular enzymes catalyze fundamental biogeochemical reactions in SOC decomposition and nutrient cycling). Cusack et al. (2010) suggests temperature-sensitivity of decomposition may be indirectly related to temperature dependence of enzyme production AND turnover. 
*	Carbon use efficiency (CUE), a measure of how well a microorganism can assimilate C into biomass relative to how much C is lost as CO2,  has been shown to vary with temperature (CITE – Lopez-Moran 2007; Price and Sowers 2004 !) with decreases in CUE at higher temperatures leading to greater CO2 production during SOM decomposition (Steinweg et al. 2008). Cite Balser and Wixon 2009 in regards to microbial-mediated effects on temp-sensitivity of C decomposition. 
*	Higher temperatures enhance desorption rates of OM with minerals (ten Hulscher and Cornelissen 1996; Kalbitz et al. 2000 – others) making more OC available in solution (Xu & Saiers, 2010; Plante et al. 2009), and thus accessible for adjacent, competent microorganisms (CITE MORE).  If we want to get technical – warmer temperatures enhance desorption of complex OC to minerals, while simultaneously providing more surface area for simple OC compounds to interact with minerals (i.e. simple C adsorps and complex C desorps in warmer temps).

@@

We are grateful to Jamie Hollingsworth for information about, and faciliating access to, the Caribou Poker Creeks Research Watershed Long-Term Ecological Research(LTER site. This research was supported by Office of Science of the U.S. Department of Energy as part of the Terrestrial Ecosystem Sciences Program. The Pacific Northwest National Laboratory is operated for DOE by Battelle Memorial Institute under contract DE-AC05-76RL01830.

@@

APS comment: So are we seeing more CH4 produced in cores that I watered? Something to note in discussion?  TODO

@@

APS comment: Did the controlled drought treatment match the 4C drought treatment in terms of rate of water loss (for both gravimetric AND volumetric)? Since the controlled drought treatment was meant to control for increased water loss with the higher temp, I want to make sure that it really was doing that. From this graph, it “looks” like it followed the same trend as the 20C drought treatment.  TODO

@@

APS comment: Was difference in cum CO2 between field moist cores and drought cores the same for both temperatures??? Granted cum Co2 was greater overall in 20C, but interesting that the difference between treatments appear equal in both temperatures. This is a bit difference from Treat et al. 2014 (granted they had permafrost/peat soils) – but Treat reported larger differences between wet/dry (aerobic/anaerobic) treatments at 20C but no differences at -0.5C, but explain this due to frozen water limiting microbial respiration.      TODO
